---
title: "Linear Models (and more!)"
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	cache = FALSE
)
```

## Outline

-   One Way ANOVA

-   Two-way ANOVA

    -   Qualitative/Qualitative

    -   Quantitative/Qualitative

-   Blocking

## Overarching Example

A plant pathology researcher is studying the effect of an experimental fungicide on the fresh weight of azaleas inoculated with a certain fungus. She has 49 potted plants of a particular azalea variety growing in a greenhouse. All plants are inoculated with a fungus. To start, she has 7 different fungicide combinations, with 7 plants per treatment.(Example from Lentner and Bishop, *Experimental Design and Analysis (2nd)*)

```{r echo = F, eval = T}
azalea <- read.csv("../data/AzaleaOneWay.csv")
```

```{r echo = TRUE, eval = F}
azalea <- read.csv("https://unl-statistics.github.io/R-workshops/r-modeling/data/AzaleaOneWay.csv")
```

```{r echo = T, eval = T}
head(azalea)
```


**Note**: As we go along, we will continue to use this scenario; we will just continue to get more specific on what the treatments are. The dataset will be updated accordingly as well.


# One-Way Analysis of Variance (ANOVA)

## One-Way ANOVA

-   Essentially an expansion of the t-test [Source](https://tjfisher19.github.io/introStatModeling/introduction-to-statistical-modeling-and-designed-experiments.html#one-way-anova)

-   Compare more than 2 levels of a treatment factor

    -   Treatment factor is your independent variable, the variable you expect to possibly cause the variation we see in the response. (i.e. the fungicide combination)

    -   Levels are the subdivisions of your overall treatment factor. (i.e. the 7 different combinations)

    -   If you only have two levels of a factor, a One-Way ANOVA would be equivalent to t-test.

-   Randomly sample your experimental units, and then randomly assign them to the different levels of the treatment factor.


## Model

$$y_{ij} = \mu + \tau_i + \varepsilon_{ij}$$

-   $y_{ij}$ is the response of the $j^{th}$ observation in the $i^{th}$ treatment level
-   $\mu$ is the overall/grand mean
-   $\tau_i$ is the effect of the $i^{th}$ treatment level
-   $\varepsilon_{ij} \sim N(0,\sigma^2)$ is the residual error

In terms of our example:

-   $y_{ij}$ is the fresh weight of the $j^{th}$ azalea in the $i^{th}$ fungicide level
-   $\mu$ is the overall/grand mean
-   $\tau_i$ is the effect of the $i^{th}$ fungicide level
-   $\varepsilon_{ij} \sim N(0,\sigma^2)$ is the residual error


## Hypotheses

$$H_0: \mu_1 = \mu_2 = ... = \mu_n$$ $$H_A: \text{at least one } \mu_i \text{ is different}$$

OR

$$H_0: \tau_1 = \tau_2 = ... = \tau_n = 0$$ $$H_A: \text{at least one } \tau_i \ne 0$$

These hypotheses are **equivalent**, just being expressed in two different ways.


## One-Way ANOVA in R

-   When running this in R, **make sure** the variables are the type you need them.
-   The following code is incorrect... why?

```{r echo=TRUE}
library(car)
#mod.fit <- lm(response ~ trt, data = data)
mod.fit <- lm(Weight ~ Trt, data = azalea)
Anova(mod.fit, type = "II")
```

-   Note the only 1 degree of freedom with treatment. We should have 6!


## One-Way ANOVA in R

-   When running this in R, **make sure** the variables are the type you need them.
-   Change `Trt` to a factor!

```{r echo=TRUE}
library(car)
#mod.fit <- lm(response ~ trt, data = data)
azalea$Trt.new <- as.factor(azalea$Trt)
mod.fit <- lm(Weight ~ Trt.new, data = azalea)
Anova(mod.fit, type = "II")
```

-   We do not have evidence of an effect of fungicide combination on the mean fresh weight of azaleas.

## But what if we did have a difference?!

-   If you reject the null hypothesis in a One-Way ANOVA, the result only tells you that **at least** one mean is different (i.e. at least one treatment level is different from the others).
-   We want to know *which* treatment levels are different!
-   Use `emmeans`

## `emmeans` results

```{r echo=TRUE}
library(emmeans)
emmeans(mod.fit, pairwise ~ Trt.new)
```

# Contrasts in R

## New Information!

Suppose you find out from the researchers that are actually only two fungicides used. Treatment 1 is a standard fungicide, Truban (**T**), and is used as a control in the experiment. Treatments 2-7 are the experimental fungicide, Nurell (**N**), that is applied at different times and rates. The researchers want to compare **T** and **N** to see if there is a difference between these two fungicide treatments.

## How do we address this new information?

-   Contrast!
-   We can write a specific test that uses our model from before to test for a difference between **T** and **N**.

| 1       | 2       | 3       | 4       | 5       | 6       | 7       |
|---------|---------|---------|---------|---------|---------|---------|
| T       | N       | N       | N       | N       | N       | N       |
| $\mu_1$ | $\mu_2$ | $\mu_3$ | $\mu_4$ | $\mu_5$ | $\mu_6$ | $\mu_7$ |

-   We would average together treatments 2-7 and compare to treatment 1.

$$H_0: \mu_1 = \frac{\mu_2 + \mu_3 + \mu_4 + \mu_5 + \mu_6 + \mu_7}{6}$$

$$H_A: \mu_1 \ne \frac{\mu_2 + \mu_3 + \mu_4 + \mu_5 + \mu_6 + \mu_7}{6}$$

## Contrasts in R

```{r}
options(width = 50)
```

-   If you care about the estimate and standard error:

```{r echo=TRUE}
library(gmodels)
fit.contrast(mod.fit, "Trt.new", c(1, -1/6, -1/6, -1/6, -1/6, -1/6, -1/6))
```


## Contrasts in R

-   If you **only care** about the test:

```{r echo=TRUE}
fit.contrast(mod.fit, "Trt.new", c(6, -1, -1, -1, -1, -1, -1))
```

-   Again, no evidence there is a difference between the mean fresh weight of **T** and **N**.


# Two-Way ANOVA

## More information!

Now that the researchers know there isn't a difference between the **T** and **N** fungicides, they want to know more about the **N** fungicide specifically. You learn that this fungicide was applied either 3 days before inoculation or 7 days after inoculation. Additionally, it is applied at either the 1 oz., 3 oz., or 5 oz. dose rates.

| N             | N             | N             | N            | N            | N            |
|------------|------------|------------|------------|------------|------------|
| 3 days before | 3 days before | 3 days before | 7 days after | 7 days after | 7 days after |
| 1 oz.         | 3 oz.         | 5 oz.         | 1 oz.        | 3 oz.        | 5 oz.        |

Now, instead of a single treatment factor, we have 2! We have time of application and rate of application. Thus, a One-Way ANOVA will no longer suffice.


## Two-Way ANOVA

-   Extension of One-Way ANOVA, we just now have two treatment factors instead of one.

-   When dealing with more than one treatment factor, our treatment design is now a *factorial*.

-   You have an $i \times j$ factorial

    -   $i$ is the number of levels of treatment 1
    -   $j$ is the number of levels of treatment 2
    -   So in our example, we have a $2 \times 3$ factorial.

-   We are interested in knowing if the two treatment factors *interact* in their effect on the response.

    -   If the response of one factor strongly depends on the level of the other factor, the two factors are said to *interact*.


## Model

$$y_{ijk} = \mu + \alpha_i + \beta_j + \alpha\beta_{ij} + \varepsilon_{ijk}$$

-   $y_{ijk}$ is the response of the $k^{th}$ observation in the $i^{th}$ treatment level of treatment 1 and the $j^{th}$ treatment level of treatment 2
-   $\mu$ is the overall/grand mean
-   $\alpha_i$ is the effect of the $i^{th}$ treatment level of treatment 1
-   $\beta_j$ is the effect of the $j^{th}$ treatment level of treatment 2
-   $\alpha\beta_{ij}$ is the interaction effect of the $i^{th}$ and $j^th$ treatment levels of treatments 1 and 2 respectively
-   $\varepsilon_{ijk} \sim N(0,\sigma^2)$ is the residual error

## Model cont.

In terms of our example:

-   $y_{ijk}$ is the fresh weight of the $k^{th}$ azalea in the $i^{th}$ treatment level of time of application and the $j^{th}$ treatment level of application rate
-   $\mu$ is the overall/grand mean
-   $\alpha_i$ is the effect of the $i^{th}$ treatment level of time of application
-   $\beta_j$ is the effect of the $j^{th}$ treatment level of application rate
-   $\alpha\beta_{ij}$ is the interaction effect of the $i^{th}$ and $j^th$ treatment levels of time of application and application rate respectively
-   $\varepsilon_{ijk} \sim N(0,\sigma^2)$ is the residual error

## Hypotheses

-   The first set of hypotheses we analyze with a Two-Way ANOVA is a test of the interaction.

$$H_0: \text{There is no interaction.}$$ $$H_A: \text{There is an interaction.}$$

-   If we **reject** the null hypothesis and conclude there is an interaction, we proceed by looking at the *simple effects*.

    -   The difference between means of two levels of one factor **at a single level of the other factor**.

-   If we **fail to reject** the null hypothesis and conclude there is no evidence of an interaction, we proceed by looking at the *main effects*.

    -   The means for each level of one factor averaged over the levels of the other.

## Two-Way ANOVA in R
```{r echo = T, eval = F}
azalea2 <- read.csv("https://unl-statistics.github.io/R-workshops/r-modeling/data/AzaleaTwoWay.csv")
```

```{r echo = F, eval = T}
azalea2 <- read.csv("../data/AzaleaTwoWay.csv")
```

```{r echo = TRUE}
azalea2$AppTime.new <- as.factor(azalea2$AppTime)
azalea2$AppRate.new <- as.factor(azalea2$AppRate)
mod.fit2 <- lm(Weight ~ AppTime.new + AppRate.new + AppTime.new:AppRate.new, data = azalea2)
Anova(mod.fit2, type = "II")
```


## Results

-   No evidence of an interaction between the time of application and application rate.
-   No evidence of a main effect of time of application.
-   We have strong evidence there is a main effect of application rate.
-   Thus, we look at the *main effect* of application rate using `emmeans`.

```{r echo=TRUE}
emmeans(mod.fit2, pairwise ~ AppRate.new)
```


## But what if there HAD been an interaction?

-   Still use the `emmeans` function!

-   In the `pairwise` part of the argument, use the vertical bar, \|, to denote which treatment factor you want to keep **constant**.

    -   Factor you want to look at differences of \| Factor you want to keep constant



## But what if there HAD been an interaction?

```{r echo=TRUE}
emmeans(mod.fit2, pairwise ~ AppRate.new|AppTime.new)
```



## But what if there HAD been an interaction?

```{r echo=TRUE}
emmeans(mod.fit2, pairwise ~ AppTime.new|AppRate.new)
```


# Factorial Models

## Another Twist!

In the previous analysis, we treated application rate as a qualitative/categorical variable, meaning the researchers are *only* interested in the levels of 1 oz., 3 oz., and 5 oz. You learn the researchers actually want to be able to predict the fresh weight of azaleas that were given **N** at the 2 oz. rate.

-   Before, we were analyzing a qualitative/qualitative factorial, where both treatment factors are qualitative.

-   Now, we want to change the analysis to a qualitative/quantitative factorial.

    -   Time of application remains a qualitative variable.

    -   Application rate is updated to a quantitative.

## Quant/Qual Factorial

-   We are now adding in *regression* to our factorial analysis.

-   Want to know if the same line can be fit for the two levels time of application

    -   In other words, does the time of application effect the prediction of fresh weight of azaleas for the rate of application?

    -   In other, other words, is there an interaction?

-   Our hypotheses are very similar to what they were earlier. We just now need to add in a test of shape for the regression lines.

    -   To determine the maximum power of your regression line, take the number of levels of your quantitative factor and subtract 1.

    -   So for our example, we have 3 levels. Thus, we can fit up to a maximum power of 2 (a quadratic).


## Quant/Qual Factorial in R

-   Need to use orthogonal contrasts in order to test for shape. Outside the scope of this workshop.
    -   These contrasts are nicely built into R.

```{r echo=TRUE}
mod.fit3 <- aov(Weight ~ AppTime.new + AppRate.new + AppTime.new:AppRate.new, data = azalea2)
summary(mod.fit3, split=list(AppRate.new = list(Linear=1, Quad=2)))
```


## Model based on our actual results

-   Need to get estimates for our coefficients. $$y = 37.4554 - 1.1786*rate + 0.4375*rate^2$$

```{r echo=TRUE}
mod.fit4 <- lm(Weight ~ AppRate + I(AppRate^2), data = azalea2)
summary(mod.fit4)
```

## If there had been an interaction?

-   For 3 days before: $y = 39.0565 - 1.3393*rate + 0.4375*rate^2$

-   For 7 days after: $y = 35.8541 - 1.0179*rate + 0.4375*rate^2$

```{r echo=TRUE}
mod.fit5 <- lm(Weight ~ AppTime.new + AppRate + I(AppRate^2) + AppTime.new:AppRate, data = azalea2)
summary(mod.fit5)
```

# Randomized Complete Block Designs

## Last Twist!

You learn the researchers actually 7 plots, and they made sure all six treatment combinations of the **N** fungicide were included. The researchers completed the experiment in a Randomized Complete Block Design! (*Note*: We are also going to go back to the assumption it is a qualitative/qualitative analysis.)

-   A *block* allows us to group experimental units together such that the variance expected within the block is a lot smaller than the variance between blocks.

    -   Think of a plot of land with a river running along one side.

-   We typically assume blocks are a random sample from the population, and thus, our sample of blocks is representative of the population of similar constructed blocks.

    -   Makes the effect of the block a *random effect*.

## RCBD Models Examples

-   One-Way ANOVA

$$y_{ij} = \mu + \tau_i + b_j + \varepsilon_{ij}$$

-   $b_j \sim N(0,\sigma_b^2)$ is the effect of the $j^{th}$ block

-   The rest of the aspects of the model are basically the same as before. Description of subscripts changes.

-   Two-Way ANOVA

$$y_{ijk} = \mu + \alpha_i + \beta_j + \alpha\beta_{ij} + b_k + \varepsilon_{ijk}$$

-   $b_k \sim N(0,\sigma_b^2)$ is the effect of the $k^{th}$ block
-   The rest of the aspects of the model are basically the same as before. Description of subscripts changes.


## RCBD Analysis in R
```{r echo = T, eval = F}
azalea3 <- read.csv("https://unl-statistics.github.io/R-workshops/r-modeling/data/AzaleaBlock.csv")
```

```{r echo = F, eval = T}
azalea3 <- read.csv("../data/AzaleaBlock.csv")
```


```{r echo = TRUE}
azalea3$AppTime.new <- as.factor(azalea3$AppTime)
azalea3$AppRate.new <- as.factor(azalea3$AppRate)
azalea3$Block.new <- as.factor(azalea3$Block)
library(lme4)
mod.fit6 <- lmer(Weight ~ AppTime.new + AppRate.new + AppTime.new:AppRate.new + (1|Block.new), data = azalea3)
Anova(mod.fit3, type = "II")
```
