---
title: "Statistical Modeling"
subtitle: "In R"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	cache = TRUE
)
```

class: center, middle

# Basic Statistical Tests


---
## Outline

+ p-values

+ confidence intervals

+ t-tests

+ chi-square tests

+ simple regression

---
## Five Step Hypothesis Testing Procedure

+ Step 1: State your null and alternate hypothesis. 

+ Step 2: Collect data. 

+ Step 3: Perform a statistical test. 

+ Step 4: Decide whether to reject or fail to reject your null hypothesis. 

+ Step 5: Present your findings.


???

+ Step 1: State your null and alternate hypothesis. 
  + (the prediction that you want to investigate)
  + The alternate hypothesis is usually your initial hypothesis that predicts a relationship between variables. The null hypothesis is a prediction of no relationship between the variables you are interested in.

+ Step 2: Collect data. 
  + For a statistical test to be valid, it is important to perform sampling and collect data in a way that is designed to test your hypothesis. If your data are not representative, then you cannot make statistical inferences about the population you are interested in.

+ Step 3: Perform a statistical test. 
  + There are a variety of statistical tests available, but they are all based on the comparison of within-group variance (how spread out the data is within a category) versus between-group variance (how different the categories are from one another).

+ Step 4: Decide whether to reject or fail to reject your null hypothesis. 
  + Based on the outcome of your statistical test, you will have to decide whether to reject or fail to reject your null hypothesis.
  + In most cases you will use the p-value generated by your statistical test to guide your decision. And in most cases, your predetermined level of significance for rejecting the null hypothesis will be 0.05 – that is, when there is a less than 5% chance that you would see these results if the null hypothesis were true.

+ Step 5: Present your findings.
  + The results of hypothesis testing will be presented in the results and discussion sections of your research paper.

---
## What is a p-value?

---
## What does a confidence interval tell me?

---
## Two-sample independent t-test

---
## Paired t-test

---
## Chi-square test

---
## Simple Regression

In a manufacturing study, the production times for 111 recent production runs were obtained. The researchers are curious if the production lot size can predict the production time in hours.

```{r, eda-example, echo = T, eval = T}
production_data <- read.csv("https://srvanderplas.github.io/rwrks/06-r-modeling/data/production.csv")
head(produciton_data) %>% kable()
```

