---
title: "Statistical Modeling"
subtitle: "In R"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	cache = TRUE
)
```

class: center, middle

# Linear Models (and more!)

---

# Outline

+ One Way ANOVA
+ Two-way ANOVA 
  + Qualitative/Qualitative
  + Quantitative/Qualitative
+ Blocking

---

# Overarching Example

A plant pathology researcher is studying the effect of an experimental fungicide on the fresh weight of azaleas inoculated with a certain fungus. She has 49 potted plants of a particular azalea variety growing in a greenhouse. All plants are inoculated with a fungus. To start, she has 7 different fungicide combinations, with 7 plants per treatment.

```{r echo = TRUE}
azalea <- read.csv("https://srvanderplas.github.io/rwrks/06-r-modeling/data/AzaleaOneWay.csv")
head(azalea)
```

**Note**: As we go along, we will continue to use this scenario; we will just continue to get more specific on what the treatments are. The dataset will be updated accordingly as well.

---

# One-Way Analysis of Variance (ANOVA)

+ Esstentially an expansion of the t-test [Source](https://tjfisher19.github.io/introStatModeling/introduction-to-statistical-modeling-and-designed-experiments.html#one-way-anova)

+ Compare more than 2 levels of a treatment factor

  - Treatment factor is your independent variable, the variable you expect to possibly cause the variation we see in the response. (i.e. the fungicide combination)
  
  - Levels are the subdivisions of your overall treatment factor. (i.e. the 7 different combinations)
  
  - If you only have two levels of a factor, a One-Way ANOVA would be equivalent to t-test.
  
+ Randomly sample your experimental units, and then randomly assign them to the different levels of the treatment factor.

---

# One-Way ANOVA Model

