---
title: "Statistical Modeling"
subtitle: "In R"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	cache = TRUE
)
```

class: center, middle

# Workshop - Bring your own data!

---
## Production

In a manufacturing study, the production times for 111 recent production runs were obtained. The researchers are curious if the production lot size can predict the production time in hours.

```{r, production-echo, purl = T, echo = T, eval = F}
production_data <- read.csv("https://srvanderplas.github.io/rwrks/06-r-modeling/data/production.csv")
head(production_data)
```

.pull-left[
```{r, production-data, purl = F, echo = F, eval = T}
production_data <- read.csv("https://srvanderplas.github.io/rwrks/06-r-modeling/data/production.csv")
head(production_data) %>% kable()
```
].pull-right[
```{r, production-eda, purl = T, echo = F, eval = T, fig.width = 4, fig.height = 4}
ggplot(data = production_data, aes(x = Lot.Size, y = Time)) +
  geom_point(shape = 1) +
  theme_bw() +
  theme(aspect.ratio = 1,
        axis.text = element_text(size = 16),
        axis.title = element_text(size = 18))
```
]

---
## Palmer Penguins (extention)

What if you wanted to fit a linear regression line between bill length and bill depth for each species?


