---
title: "Your Turn Solutions"
output: html_document
---

## Motivating Example

Try playing with chunks of code from this session to further investigate the tips data:

1. Get a summary of the daily shedding values (use the shed data set)

```{r}

#shed <- read_csv("https://srvanderplas.github.io/rwrks/01-r-intro/data/daily_shedding.csv")

shed <- read.csv("../data/daily_shedding.csv")


summary(shed$daily_shedding)
```

2. Make side by side boxplots of final weight gain by treatment group (use the final_shed data set)

```{r message=FALSE}
library(tidyverse)

final_shed <- shed %>% 
  group_by(pignum) %>% 
  mutate(gain = pig_weight[time_point == 21] - pig_weight[time_point == 0]) %>% filter(time_point == 21)  %>% ungroup() %>% select(-c(4:9))

ggplot(final_shed) + 
  geom_boxplot( aes(treatment, gain, fill = treatment))
```

3. Compute a wilcox test for control vs. the "Bglu" treatment group

```{r}
wilcox.test(total_shedding ~ treatment, data = final_shed,
            subset = treatment %in% c("control", "Bglu"))
```

## Basics

Using the R Reference Card (and the Help pages, if needed), do the following:

1. Find out how many rows and columns the `iris' data set has.  Figure out at least 2 ways to do this.
*Hint: "Variable Information" section on the first page of the reference card!*

```{r}
nrow(iris)
ncol(iris)
dim(iris)
```

2. Use the `rep` function to construct the following vector:  1 1 2 2 3 3 4 4 5 5
*Hint: "Data Creation" section of the reference card*

```{r}
rep(1:5, each = 2)
```

3. Use `rep` to construct this vector: 1 2 3 4 5 1 2 3 4 5 1 2 3 4 5

```{r}
rep(1:5, times = 3)
```

1. Find out how many pigs had a total shedding value of less than 30 log10 CFUs.
*Hint: if you use the `sum` function on a logical vector, it'll return how many TRUEs are in the vector:*
```{r}
sum(c(TRUE, TRUE, FALSE, TRUE, FALSE))
```

```{r}
sum(shed$total_shedding < .30)
```

2. **More Challenging**: Calculate the sum of the total shedding log10 CFUs of all pigs with a total shedding value of less than 30 log10 CFUs

```{r}
sum(shed$total_shedding[shed$total_shedding < 30])
```

1. Which pigs have a shedding value less than or equal to 30 OR is in the Acid treatment group?

```{r}
shedding <- final_shed$total_shedding
treatment <- final_shed$treatment
id <- (shedding <= 30 | treatment == "Acid")
final_shed[id,]
```

## Data Structures

1. In row_matrix, change 'byrow = FALSE' to see what happens

```{r}
matrix(
  c(1, 2, 3, 4, 5, 6, 7, 8, 9),
  nrow = 3,  
  ncol = 3,        
  byrow = FALSE        
)
```


2. Recreate this matrix.

```{r}
matrix(c(2,4,6,8,10,12), nrow=3, ncol = 2, byrow = FALSE)
```

1. Make a data frame with column 1: 1,2,3,4,5,6 and column 2: a,b,a,b,a,b

```{r}
mydf <- data.frame(col1 = 1:6, col2 = rep(c("a", "b"), times = 3))
```

2. Select only rows with value "a" in column 2 using logical vector

```{r}
mydf[mydf$col2 == "a",]
```

3.
```{r}
names(mydf) <- c("Bulldogs", "Tigers")
```


4. `mtcars` is a built in data set like `iris`: Extract the 4th row of the `mtcars` data.

```{r}
mtcars[4,]
```


5. Create another column in the flower data frame, that is the sum of Sepal Width, Sepal Length, Petal Width and Petal Length.

```{r}
flower <- iris

total <- flower$Sepal.Length + flower$Sepal.Width + flower$Petal.Length + flower$Petal.Width

flower2 <- data.frame(flower, total)

head(flower2)

```

6. Step Further. Create a data frame of observations where Sepal Length is greater than 6. Then create another column in the this data frame, that is the sum of Sepal Width and Sepal Length


```{r}
flower_s <- flower[flower$Sepal.Length>6, ]

total_s <- flower_s$Sepal.Length + flower_s$Sepal.Width

head(data.frame(flower_s, total_s))

```




1. Create a list containing a vector and a 2x3 data frame

```{r}
mylist <- list(vec = 1:6, df = data.frame(x = 1:2, y = 3:4, z = 5:6))
```

2. Use indexing to select the data frame from your list

```{r}
mylist[[2]]
```

3. Use further indexing to select the first row from the data frame in your list

```{r}
mylist[[2]][1,]
```

1. View the top 8 rows of mtcars data

```{r}
head(mtcars, n = 8)
```

2. What type of object is the mtcars data set?

```{r}
str(mtcars)
```

3. How many rows are in iris data set? (try finding this using dim or indexing + length)

```{r}
dim(iris)
```

4. Summarize the values in each column in iris data set

```{r}
summary(iris)
```

## Packages and Programming

1. Create a function that takes numeric input and provides the mean and a 95% confidence interval for the mean for the data (the t.test function could be useful)

```{r}
mean_and_ci <- function(x) {
    themean <- mean(x)
    theci <- t.test(x)$conf.int
    
    return(list(mean = themean, ci = theci))
}
```

2. Add checks to your function to make sure the data is either numeric or logical.  If it is logical convert it to numeric.

```{r}
mean_and_ci <- function(x) {
    if (!is.numeric(x) && !is.logical(x)) stop("Need logical or numeric data")
    
    x <- as.numeric(x)
    
    themean <- mean(x)
    theci <- t.test(x)$conf.int
    
    return(list(mean = themean, ci = theci))
}
```

3. Loop over the columns of the diamonds data set and apply your function to all of the numeric columns.

```{r}
for (i in colnames(diamonds)) {
  if (is.numeric(diamonds[,i,drop = TRUE])) {print(mean_and_ci(diamonds[,i,drop = TRUE]))}
  else {print("nope")}
}
```

