---
title: "Statistical Modeling in R (Planning)"
author: "Emily Robinson, Denise Bradford, & Rachel Rogers"
date: "Spring 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

Learn how to fit models, test hypotheses, create model output tables and plots, and more! In the morning, we will cover model-fitting basics (linear regression, mixed models), and in the afternoon we will look into more advanced types of models (generalized additive models, generalized linear mixed models, etc.)

## Target Audience

We assume the participants have taken a class where they learned ANOVA, regression, some experimental design, and maybe generalized linear models (e.g. 801 and 802). The goal is to teach them how to implement the models in R and extract and display the output, not to teach them what the models are doing.

## Packages

```{r packages, eval = F}
library(lme4) # lm, glm, lmer, glmer glm.nb glmer.nb
library(nlme) # repeated measures
library(nls)  # nonlinear regression?
library(mgcv) # generalized additive mixed models

library(emmeans) # obtains estimates
library(broom) # cleans up model output
```

## Topics

+ Randomization / Experimental Design
+ ANOVA
+ ANCOVA
+ MANOVA??
+ Linear Regression, Multiple Linear Regression
+ Generalized Models
+ Mixed Models
+ Dimension Reduction / Variable Selection
  + Stepwise selection, PCA, NMDS, LASSO, varImp, etc.
+ What to spot - how to determine your test?
  
## Order
+ t.test, chi.sq test
+ lm (ANOVA, ANCOVA, linear regression, multiple linear regression (stepwise selection, etc.))
  + crd experimental
  + one way
  + two way
  + quant and qual
+ lmer
  + rcbd, split plot
+ glm
  + change response type
+ glmer

## Examples
+ Pick one or two scenarios and modify for the different types of tests.
