---
title: "Web APIs in R"
subtitle: "Using jsonlite"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

# Web APIs

## Web APIs {.build} 

- [Server-side Web APIs](https://en.wikipedia.org/wiki/Web_API#Server-side) (Application Programming Interfaces) are a popular way to provide easy access to data and other services. 
- If you (the client) want data from a server, you typically need one HTTP verb -- `GET`.

```{r}
library(httr)
sam <- GET("https://api.github.com/users/sctyner")
content(sam)[c("name", "company")]
```

- Other HTTP verbs -- `POST`, `PUT`, `DELETE`, etc...

## Request/response model {.build}

- When you (the client) _requests_ a resource from the server. The server _responds_ with a bunch of additional information.

```{r}
sam$header[1:3]
```

- Nowadays content-type is usually XML or JSON (HTML is great for _sharing content_ between _people_, but it isn't great for _exchanging data_ between _machines_.)

# Non-HTML Data Formats

## What is XML?

XML is a markup language that looks very similar to HTML.

```xml
<mariokart>
  <driver name="Bowser" occupation="Koopa">
    <vehicle speed="55" weight="25"> Wario Bike </vehicle>
    <vehicle speed="40" weight="67"> Piranha Prowler </vehicle>
  </driver>
  <driver name="Peach" occupation="Princess">
    <vehicle speed="54" weight="29"> Royal Racer </vehicle>
    <vehicle speed="50" weight="34"> Wild Wing </vehicle>
  </driver>
</mariokart>
```

- This example shows that XML can (and is) used to store inherently tabular data ([thanks Jeroen Ooms](http://arxiv.org/pdf/1403.2805v1.pdf))
- What is are the observational units here? How many observations in total?
- 2 units and 6 total observations (4 vehicles and 2 drivers).

## XML2R {.build}

[XML2R](https://github.com/cpsievert/XML2R) is a framework to simplify acquistion of tabular/relational XML.

```{r, echo=FALSE, message = FALSE, warning=FALSE}
# hopefully no one is watching
# devtools::install_github("cpsievert/XML2R")
library(XML2R)

#obs <- read_xml("https://gist.githubusercontent.com/cpsievert/85e340814cb855a60dc4/raw/651b7626e34751c7485cff2d7ea3ea66413609b8/mariokart.xml")
obs <- XML2Obs("https://gist.githubusercontent.com/cpsievert/85e340814cb855a60dc4/raw/651b7626e34751c7485cff2d7ea3ea66413609b8/mariokart.xml", quiet = TRUE)
data.frame(obs)
table(names(obs))
```

* **XML2R** coerces XML into a *flat* list of observations.
* The list names track the "observational unit".
* The list values track the "observational attributes".

---

```{r}
obs # named list of observations
```

---

```{r}
collapse_obs(obs) # group into table(s) by observational name/unit
```

- What information have I lost?
- I can't map vehicles to the drivers!

---

```{r}
library(dplyr)
obs <- add_key(obs, parent = "mariokart//driver", recycle = "name")
collapse_obs(obs)
```

---

Now (if I want) I can merge the tables into a single table...

```{r}
tabs <- collapse_obs(obs)
left_join(as.data.frame(tabs[[1]]), as.data.frame(tabs[[2]])) 
```

## What about JSON? {.build}

- JSON is _the_ format for data on the web.
- JavaScript Object Notation (JSON) is comprised of two components:
    * arrays => [value1, value2]
    * objects => {"key1": value1, "key2": [value2, value3]} 
- [jsonlite](http://cran.r-project.org/web/packages/jsonlite/index.html) is the preferred R package for parsing JSON (it's even used by Shiny!)

## Back to Mariokart {.smaller}

```json
[
    {
        "driver": "Bowser",
        "occupation": "Koopa",
        "vehicles": [
            {
                "model": "Wario Bike",
                "speed": 55,
                "weight": 25
            },
            {
                "model": "Piranha Prowler",
                "speed": 40,
                "weight": 67
            }
        ]
    },
    {
        "driver": "Peach",
        "occupation": "Princess",
        "vehicles": [
            {
                "model": "Royal Racer",
                "speed": 54,
                "weight": 29
            },
            {
                "model": "Wild Wing",
                "speed": 50,
                "weight": 34
            }
        ]
    }
]
```


---

```{r}
library(jsonlite)
mario <- fromJSON("http://bit.ly/mario-json")
str(mario) 
```

---

```{r}
mario$driver
mario$vehicles
```

How do we get two tables (with a common id) like the XML example?

```{r}
vehicles <- rbind(mario$vehicles[[1]], mario$vehicles[[2]])
vehicles <- cbind(driver = mario$driver, vehicles)
```

## Your Turn {data-background=#527a7a}

1. Get the json data for our R workshop GitHub commit history:

```{r}
workshop_commits_raw <- fromJSON("https://api.github.com/repos/heike/rwrks/commits")
```

2. Find the table of commits contained in this list. Hint: It's all about the `$`

3. Plot the total number of commits (number of rows) by user as a bar chart

